---
title: "QUMIs"
author: "Will Townes"
output: html_document
---

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(tidyverse); theme_set(theme_bw())
library(ggridges)
fp<-file.path
plt_dir<-fp("./qumi/results/fig")
if(!dir.exists(plt_dir)){ dir.create(plt_dir,recursive=TRUE) }
```

## Training Data Plots

load MLEs and BIC values for the seven training datasets for all three models.

```{r}
klein_celltypes<-SingleCellExperiment::colData(readRDS("./real/klein_2015/data/01_sce_all_genes_all_cells.rds"))$cell_type1
klein_groups<-as.character(klein_celltypes)
klein_groups[klein_groups %in% c("d0","d7")]<-"d0d7"
klein_groups[klein_groups %in% c("d2","d4")]<-"d2d4"

dats<-c("cao_2017","clark_2019","grun_2016","klein_2015","schiebinger_2019","zeisel_2015","zhang_2019")
mods<-c("nb","poilog","poilomax")
res<-list()
for(m in mods){
  res[[m]]<-list()
  for(d in dats){
    df<-read.table(fp("./real",d,"results",paste0("mle_",m,".txt")),header=TRUE)
    df$dataset<-if(d=="klein_2015"){ paste(d,klein_groups,sep="_") } else { d }
    df$model<-m
    df$cell_id<-rownames(df)
    res[[m]][[d]]<-df
  }
  res[[m]]<-do.call(rbind,res[[m]])
  res[[m]]$dataset<-factor(res[[m]]$dataset,levels=sort(unique(res[[m]]$dataset),decreasing=TRUE))
  rownames(res[[m]])<-NULL
}
#extract BIC and log likelihoods
bic_df<-lapply(res,function(x){x[,c("dataset","cell_id","model","loglik","bic")]})
bic_df<-do.call(rbind,bic_df)
rownames(bic_df)<-NULL
#extract MLEs
res2<-res
res2[["nb"]]<-res[["nb"]][,c("dataset","cell_id","model","size")]
res2[["poilog"]]<-res[["poilog"]][,c("dataset","cell_id","model","sig")]
res2[["poilomax"]]<-res[["poilomax"]][,c("dataset","cell_id","model","tail")]
#for(m in mods){
#  colnames(res2[[m]])[4]<-"shape"
#}
#mle_df<-do.call(rbind,res2)
#rownames(mle_df)<-NULL
```

Make plots of MLEs

```{r}
ggplot(res2[["nb"]],aes(x=size,y=dataset,fill=dataset))+geom_density_ridges(scale=1.8,alpha=.9,rel_min_height=.0001)+theme(legend.position="none")+xlab("shape")+scale_x_log10()
ggsave(fp(plt_dir,"nb_mles.pdf"),width=3,height=2.7)

#exclude shape<.1, only small number of outlier cells this low 
rplg<-subset(res2[["poilog"]],sig>.1)
ggplot(rplg,aes(x=sig,y=dataset,fill=dataset))+geom_density_ridges(scale=1.8,alpha=.9,rel_min_height=.0001)+theme(legend.position="none")+xlab("shape")+ylab(NULL)+scale_x_log10()
ggsave(fp(plt_dir,"poilog_mles.pdf"),width=3,height=2.7)
ggsave(fp(plt_dir,"poilog_mles_big.pdf"),width=6,height=4)

ggplot(res2[["poilomax"]],aes(x=tail,y=dataset,fill=dataset))+geom_density_ridges(scale=1.8,alpha=.9,rel_min_height=.0001)+theme(legend.position="none")+scale_x_log10(limits=c(min(res2[["poilomax"]]$tail),10))+xlab("shape")+ylab(NULL)
ggsave(fp(plt_dir,"poilomax_mles.pdf"),width=3,height=2.7)
```

Does Poisson-lognormal shape parameter correlate with total counts or detection rate?

```{r}
#dats<-c("cao_2017","clark_2019","grun_2016","klein_2015","schiebinger_2019","zeisel_2015","zhang_2019")
pd<-res2[["poilog"]]
pd$model<-NULL
pd$dataset<-as.character(pd$dataset)
pd$dataset[grepl("klein_2015_",pd$dataset,fixed=TRUE)]<-"klein_2015"
pd$dataset<-factor(pd$dataset)
f<-function(d){
  if(d=="clark_2019"){
    scefile<-"02_sce_subset_p14.rds"
  } else {
    scefile<-"01_sce_all_genes_all_cells.rds"
  }
  m<-counts(readRDS(fp("real",d,"data",scefile)))
  tot<-colSums(m)
  pz<-1-colMeans(m>0)
  tot_gene<-colSums(m>0)
  data.frame(dataset=d,cell_id=colnames(m),total_umi=tot,zero_frac=pz,total_genes=tot_gene)
}
res<-do.call(rbind,lapply(dats,f))
pd2<-merge(pd,res,by=c("dataset","cell_id"),all=FALSE)
ggplot(pd2,aes(x=total_umi,y=sig))+geom_point()+facet_wrap(~dataset,scales="free")+scale_x_log10()
ggplot(pd2,aes(x=zero_frac,y=sig))+geom_point()+facet_wrap(~dataset,scales="free")
ggplot(pd2,aes(x=total_genes,y=sig))+geom_point()+facet_wrap(~dataset,scales="free")
knitr::kable(pd2 %>% group_by(dataset) %>% subset(!is.na(sig)) %>% summarize(corr=cor(sig,total_genes)) %>% arrange(corr))
```

Make plots of BIC

```{r}
bic_df$dataset<-as.character(bic_df$dataset)
bic_df$dataset[grep("klein_2015",bic_df$dataset)]<-"klein_2015"
ggplot(bic_df,aes(x=model,fill=model,y=bic))+geom_boxplot()+facet_wrap(~dataset,scales="free",nrow=3)+scale_y_log10()+theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),legend.position=c(.5,.15))+ylab("Bayesian information criterion (BIC)")
ggsave(fp(plt_dir,"bic_boxplot.pdf"),width=6,height=6)
```

Load and plot results from comparisons based on simulated maxima

```{r}
dats<-c("cao_2017","clark_2019","grun_2016","klein_2015","schiebinger_2019","zeisel_2015","zhang_2019")
f<-function(d){
  x<-read.table(fp("real",d,"results/compare_maxima.txt"),header=TRUE)
  x$dataset<-d
  x
}
pd<-do.call(rbind,lapply(dats,f))
pd$log_err<-log(pd$sim_max)-log(pd$true_max)
ggplot(pd,aes(x=log_err))+geom_histogram(bins=50)+facet_grid(dataset~method,scales="free")+geom_vline(xintercept=0,col="red")+xlab("log ratio of simulated versus observed maxima")+ylab("number of cells in bin")
ggsave(fp(plt_dir,"cpmax_hist.pdf"),width=6,height=9)
```

## Test Data Plots

Load distance scores

```{r}
dats<-c("macosko_2015","tung_2016","zheng_2017_monocytes")
res<-list()
for(d in dats){
  res[[d]]<-read.table(fp("./real",d,"results/qumi_distance.txt"),header=TRUE)
  res[[d]]$dataset<-d
}
res0<-res #cache for use by later code cells
```

```{r}
res<-res0
keep<-c("dataset","cell","read_counts","census_counts")
res$macosko_2015<-res$macosko_2015[,c(keep,"qumi_poilog_1.9","qumi_poilog_2")]
res$tung_2016<-res$tung_2016[,c(keep,"qumi_poilog_2.4","qumi_poilog_2")]
res$zheng_2017_monocytes<-res$zheng_2017_monocytes[,c(keep,"qumi_poilog_2.4","qumi_poilog_2")]
for(d in dats){
  colnames(res[[d]])[5:6]<-paste0("qumi_",c("custom","default"))
}
res<-do.call(rbind,res)
rownames(res)<-NULL

pd<-gather(res,key="method",value="distance",-c(dataset,cell))
pd2<-pd %>% group_by(method) %>% summarize(med_dist=median(distance,na.rm=TRUE))
o<-order(pd2$med_dist,decreasing=FALSE)
pd$method<-factor(pd$method,levels=pd2$method[o])
ggplot(pd,aes(x=method,y=distance,fill=method,colour=method))+geom_boxplot()+scale_y_log10()+ylab("distance from UMI counts")+facet_wrap(~dataset,nrow=1,scales="free")+theme(legend.position="none")+theme(axis.text.x = element_text(angle = 20, hjust=1)) #+coord_flip()
ggsave(fp(plt_dir,"dist_comp.pdf"),width=7,height=3)
```

sensitivity analysis: examine effect of a grid of values for poilog shape param

```{r}
res<-res0
keep<-c("dataset","cell","read_counts","census_counts",paste0("qumi_poilog_",1:3))
for(d in dats){
  res[[d]]<-res[[d]][,keep]
}
res<-do.call(rbind,res)
rownames(res)<-NULL
cn<-colnames(res)
cn<-sub("qumi_poilog_","qumi",cn,fixed=TRUE)
cn<-sub("census_counts","census",cn,fixed=TRUE)
cn<-sub("read_counts","reads",cn,fixed=TRUE)
colnames(res)<-cn

pd<-gather(res,key="method",value="distance",-c(dataset,cell))
pd2<-pd %>% group_by(method) %>% summarize(med_dist=median(distance,na.rm=TRUE))
o<-order(pd2$med_dist,decreasing=FALSE)
pd$method<-factor(pd$method,levels=pd2$method[o])
ggplot(pd,aes(x=method,y=distance,fill=method,colour=method))+geom_boxplot()+scale_y_log10()+ylab("distance from UMI counts")+facet_wrap(~dataset,nrow=1,scales="free")+theme(legend.position="none")+theme(axis.text.x = element_text(angle = 20, hjust=1)) #+coord_flip()
ggsave(fp(plt_dir,"dist_comp_sensitivity.pdf"),width=7,height=3)
```

are distance scores affected by total read counts?

```{r}
dats<-c("macosko_2015","tung_2016","zheng_2017_monocytes")
sce<-list()
sce[["macosko_2015"]]<-readRDS("./real/macosko_2015/data/02_sce_subset_r6.rds")
sce[["tung_2016"]]<-readRDS("./real/tung_2016/data/01_sce_all_genes_all_cells.rds")
sce[["zheng_2017_monocytes"]]<-readRDS("./real/zheng_2017_monocytes/data/01_sce_all_genes_all_cells.rds")
sce2rc<-function(d){
  rc<-Matrix::colSums(SummarizedExperiment::assay(sce[[d]],"read_counts"))
  data.frame(dataset=d,cell=names(rc),total_read_counts=rc)
}
rc<-do.call(rbind,lapply(dats,sce2rc))
pd4<-merge(pd3,rc)
ggplot(pd4,aes(x=total_read_counts,y=distance,colour=method))+geom_point()+facet_wrap(~dataset,scales="free")+scale_y_log10()+scale_x_log10()
```

for all methods, the distance increases as total read count of a cell increases.

### MA plots

```{r}
res1<-read.table(fp("./real/tung_2016/results/qumi_maplot.txt"),header=TRUE)
res2<-read.table(fp("./real/zheng_2017_monocytes/results/qumi_maplot.txt"),header=TRUE)
res1$dataset<-"tung_2016"
res2$dataset<-"zheng_2017"
res<-rbind(res1,res2)
res$method<-factor(res$method,levels=c("read_counts","scran_counts","census_counts","qumi_counts"))

ggplot(res,aes(x=mean_a,y=mean_m))+geom_hex(color="black",fill="black",bins=200,show.legend=FALSE)+facet_grid(dataset~method,scales="free_x")+xlab("average log expression (A)")+ylab("difference in log expression versus UMI counts (M)")+geom_hline(yintercept=0,colour="red")
ggsave(fp(plt_dir,"maplot_mean.pdf"),width=6,height=4)

ggplot(res,aes(x=cv_a,y=cv_m))+geom_hex(color="black",fill="black",bins=200,show.legend=FALSE)+facet_grid(dataset~method,scales="free_x")+xlab("average log CV (A)")+ylab("difference in log CV versus UMI counts (M)")+geom_hline(yintercept=0,colour="red")
ggsave(fp(plt_dir,"maplot_cv.pdf"),width=6,height=4)
```
