---
title: "Quasi-UMIs for Macosko 2015"
author: "Will Townes"
output: html_document
---

Exploring reads vs UMIs with goal of normalizing read counts to quasi-UMIs

* Estimate discrete power law distribution for many UMI samples
* Quantile normalize read counts to quasi UMI counts
* Compare qumi distribution to actual UMI distribution

```{r}
library(tidyverse); theme_set(theme_bw())
suppressPackageStartupMessages(library(SingleCellExperiment))
source("./algs/quminorm.R") #also loads nblomax.R,poilog.R
source("./util/misc.R") #splitMat function
fp<-file.path
bp<-"./real/macosko_2015"

plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)){ dir.create(plt_dir,recursive=TRUE) }
```

data loading

```{r}
sce<-readRDS(fp(bp,"data/02_sce_subset_r6.rds"))
m<-counts(sce)
rc<-assay(sce,"read_counts")
```

PMF of random cell, compare poisson-lomax with poisson-lognormal

```{r}
#i<-1
i<-sample.int(ncol(m),size=1)
x<-m[,i]
(th<-poilog_mle(x))
(th2<-plomax_mle(x,quadpts=1000))
(th3<-nb_mle(x))
lpmf(x,cex=2)
llcurve_poilog(max(x),th,col="blue",lwd=2)
llcurve_lomax(max(x),th2,lik="poi",col="green",lty=2,lwd=2,q=5000)
llcurve_nb(max(x),th3,col="red",lty=3,lwd=2)
legend("topright",c("poisson-lognormal","poisson-lomax","negative binomial"),col=c("blue","green","red"),lty=c(1,2,3),lwd=2)
```

make lpmf plots for all cells

```{r}
f<-function(t){
  df<-lpmf(m[,t],doplot=FALSE)
  df$id<-t
  df
}
res<-do.call(rbind,lapply(seq.int(ncol(m)),f))
ggplot(res,aes(x=x,y=y,group=id))+geom_line(alpha=.1)+xlab("log(1+x)")+ylab("log(density)")
ggsave(fp(plt_dir,"lpmf_allcells_macosko.pdf"),width=6,height=4)
```

### Apply QUMI to entire data matrix 

First we create a configuration file (JSON) to keep track of all the shape parameters for each model. Note that for Poisson-lognormal, 1.9 is the median of the iPSC-specific MLE distribution in the *clark_2019* training dataset.

```{r}
shp<-list(nb=.1,poilog=c(1,2,1.9,3),plomax=1)
write(jsonlite::toJSON(shp,pretty=TRUE),file=fp(bp,"quminorm_script_config.json"))
```

```{bash}
Rscript ./real/util/quminorm_script.R ./real/macosko_2015 data/02_sce_subset_r6.rds data/03_sce_qumi_r6.rds
```
or 
```{bash}
sbatch -J macosko ./real/util/quminorm_script.slurm ./real/macosko_2015 data/02_sce_subset_r6.rds data/03_sce_qumi_r6.rds
```

compare distances between normalizations for all cells. Using the UMI counts as a gold standard, we compute the euclidean distance between nonzero normalized counts and the UMI counts on the log scale for each normalization.

```{r}
sce2<-readRDS(fp(bp,"data/03_sce_qumi_r6.rds"))
system.time(res<-distance_compare(sce2))
write.table(res,file=fp(bp,"results/qumi_distance.txt"),quote=FALSE,row.names=FALSE)
```
or
```{bash}
Rscript ./real/util/qumi_dist_compare_script.R ./real/macosko_2015/data/03_sce_qumi_r6.rds ./real/macosko_2015/results/qumi_distance.txt
```
or
```{bash}
sbatch -J macosko_dist_comp ./real/util/qumi_dist_compare_script.slurm ./real/macosko_2015/data/03_sce_qumi_r6.rds ./real/macosko_2015/results/qumi_distance.txt
```

### see if quasi-UMI improves over other normalizations

graphical comparison of methods. Qumis are nearly always closer to the true UMI counts than read counts or CPM.

```{r}
res<-read.table(fp(bp,"results/qumi_distance.txt"),header=TRUE,stringsAsFactors=FALSE)
pd<-tidyr::gather(res,key="method",value="distance",-cell)
pd<-subset(pd,method!="counts")
pd2<-pd %>% group_by(method) %>% summarize(med_dist=median(distance,na.rm=TRUE))
o<-order(pd2$med_dist)
pd$method<-factor(pd$method,levels=pd2$method[o])
ggplot(pd,aes(x=method,y=distance))+geom_boxplot()+scale_y_log10()+theme(axis.text.x = element_text(angle = 15, hjust=1))+ylab("distance from UMI counts")
ggsave(fp(plt_dir,"dist_comp_macosko.pdf"),width=6,height=4)
```

### Filtering genes based on percentage expressed

What if we prefilter genes based on a percentage threshold instead of a fixed threshold? Does splitting the data into 10 smaller batches affect the quminorm results? 

```{r}
filter_genes_pct<-function(sce,thresh=.01,assay_name="counts"){
  #removes all genes that are not expressed in at least .01*n cells
  #.01= (1 percent) and n=ncol(m)=number of total cells in the data matrix
  #returns a SingleCellExperiment whose rows are the subset of rows of m satisfying the criterion.
  #columns of the matrix with all zeros are also removed.
  m<-assay(sce,assay_name)
  gg<-rowSums(m>0) > (thresh*ncol(m))
  sce<-sce[gg,]
  gc<-colSums(m)>0
  sce[,gc]
}

qumifunc<-function(sce){
  quminorm_matrix(assay(sce,"read_counts"),2.0,lik="poilog")
}

sce_pct1<-filter_genes_pct(sce,thresh=.01)
system.time(assay(sce_pct1,"qumi_poilog")<-qumifunc(sce_pct1)) #17 mins

x<-splitMat(sce,10)
res<-list()
res_pct1<-list()
for(i in seq_along(x)){
  sce_i<-sce[,x[[i]]]
  assay(sce_i,"qumi_poilog")<-qumifunc(sce_i)
  res[[i]]<-sce_i
  sce_pct1_i<-filter_genes_pct(sce_i,thresh=.01)
  assay(sce_pct1_i,"qumi_poilog")<-qumifunc(sce_pct1_i)
  res_pct1[[i]]<-sce_pct1_i
}
#saveRDS(sce_pct1,file=fp(bp,"data/04_sce_qumi_r6_pct1.rds"))
#saveRDS(res,file=fp(bp,"data/04_split_sce_qumi_r6.rds"))
#saveRDS(res_pct1,file=fp(bp,"data/04_split_sce_qumi_r6_pct1.rds"))
d1<-distance_compare(sce_pct1)
d2<-do.call(rbind,lapply(res,distance_compare))
d3<-do.call(rbind,lapply(res_pct1,distance_compare))
d0<-read.table(fp(bp,"results/qumi_distance.txt"),header=TRUE)
d0<-d0[,colnames(d1)]
d0$thresh<-"fixed"; d0$is_split<-FALSE
d1$thresh<-"percent"; d1$is_split<-FALSE
d2$thresh<-"fixed"; d2$is_split<-TRUE
d3$thresh<-"percent"; d3$is_split<-TRUE
res<-rbind(d0,d1,d2,d3)
res$counts<-NULL
write.table(res,file=fp(bp,"results/qumi_distance_qc_compare.txt"),quote=FALSE,row.names=FALSE)
```
```{r}
res<-read.table(fp(bp,"results/qumi_distance_qc_compare.txt"),header=TRUE,stringsAsFactors=FALSE)
ggplot(res,aes(x=thresh,y=qumi_poilog,fill=is_split))+geom_boxplot()+scale_y_log10()+ylab("distance from UMI")+xlab("gene filtering method")
res2<-res %>% group_by(cell,thresh) %>% summarize(chg=qumi_poilog[!is_split]-qumi_poilog[is_split])
ggplot(res2,aes(x=thresh,y=chg,fill=thresh))+geom_boxplot()+xlab("gene filtering method")+ylab("change in dist from UMI merged vs split")
```

```{r}
#Poisson-lomax tends to overestimate
plot(log1p(umi),log1p(xqumi),main="Poisson-Lomax Quasi-UMIs")
abline(0,1)
#neg binom tends to underestimate
plot(log1p(umi),log1p(xqumi3),main="Negative Binomial Quasi-UMIs")
abline(0,1)
#Poisson-lognormal does well
plot(log1p(umi),log1p(xqumi2),main="Poisson-Lognormal Quasi-UMIs")
abline(0,1)
#census counts
plot(log1p(umi),log1p(cc),main="Census Counts")
abline(0,1)
```

