---
title: "Quasi-UMIs for Klein 2015 (indrops)"
author: "Will Townes"
output: html_document
---

Estimate poisson-lomax, poisson-lognormal, and negative binomial (poisson-gamma) parameters for each cell

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
source("./algs/quminorm.R") #also loads nblomax.R
source("./algs/poilog.R")
fp<-file.path
bp<-"./real/klein_2015"
plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)) dir.create(plt_dir,recursive=TRUE)
```

data loading

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-counts(sce)
cm<-colData(sce)
```

PMF of random cell, compare poisson-lomax with poisson-lognormal

```{r}
i<-1146
colnames(m)[i] #cell ID="cell1146"
#set.seed(222)
#i<-sample.int(ncol(m),size=1)
x<-m[,i]
(th<-poilog_mle(x))
(th2<-plomax_mle(x,quadpts=1000))
res<-lpmf(x,doplot=FALSE)
xmax<-expm1(6.5)

fpoilog<-function(t){
  sads::dpoilog(floor(exp(t)),mu=th[1],sig=th[2],log=TRUE)
}
fplomax<-function(t){
  dplomax(floor(exp(t)),tail=th2[1],scale=th2[2],log=TRUE,quadpts=5000)
}

pdf(file=fp(plt_dir,"lpmf_llcurve_klein.pdf"),width=5,height=5)
lpmf_xtra(res,xlab="UMI count",ylab="log(density)",connect=FALSE,ylim=c(-18,0),xlim=c(-log(2),log(xmax)),cex=1.2)
curve(fpoilog,from=log(1),to=log(xmax),add=TRUE,col="blue",lwd=2)
curve(fplomax,from=log(1),to=log(xmax),add=TRUE,col="green",lwd=2,lty=2)
points(rep(-log(2),2),c(fpoilog(-Inf),fplomax(-Inf)),col=c("blue","green"),pch=c(1,19),cex=c(.8,.3))
legend("topright",c("Poisson-lognormal","Poisson-Lomax"),col=c("blue","green"),lty=c(1,2),lwd=2)
dev.off()
```

Compute MLEs for all cells `time Rscript ./real/util/mles_qumi.R ./real/klein_2015` or `sbatch -J klein ./real/util/mles_qumi.slurm ./real/klein_2015`

```{r}
nb<-read.table(fp(bp,"data/mle_nb.txt"),header=TRUE)
lx<-read.table(fp(bp,"data/mle_poilomax.txt"),header=TRUE)
ln<-read.table(fp(bp,"data/mle_poilog.txt"),header=TRUE)
good<-!is.na(lx$tail) & !is.na(ln$mu) & !is.na(nb$size)
table(good) #all good
#when is poilog best?
mean(ln$bic<pmin(lx$bic,nb$bic))
#when is poisson-lomax best?
mean(lx$bic<pmin(ln$bic,nb$bic))
#nb best?
mean(nb$bic<pmin(ln$bic,lx$bic))
```

Based on BIC, in 60% of cells, Poisson-lognormal fits best. In 40% of cells Poisson-lomax fits best. In 0% of cells, negative binomial fits best.

Comparing MLEs of scale parameter to MOM of scale param with fixed shape

```{r}
lpz<-log(1-colMeans(m>0))
scale_mom<-poilog_pzero2mu(lpz,sig=2.0)
pd<-data.frame(scale_mom=scale_mom,celltype=cm$cell_type1,shape_mle=ln$sig,scale_mle=ln$mu)
ggplot(pd,aes(x=scale_mle,y=scale_mom,colour=shape_mle))+geom_point()+geom_abline(intercept=0,slope=1)+scale_color_gradient2(low="orange",mid="blue",high="red",midpoint = 2.0)+xlab("scale parameter MLE")+ylab("scale parameter MOM estimate")
ggsave(fp(plt_dir,"klein_mom_vs_mle.pdf"),width=6,height=4)
```

Comparing fitted MLEs to summary statistics

```{r}
cm<-colData(sce)
z<-log10(colSums(m))
hist(ln$sig,breaks=100)
boxplot(ln$sig~cm$cell_type1)
boxplot(z~cm$cell_type1)
plot(z,ln$sig,col=cm$cell_type1)
plot(z,ln$mu)
plot(z,ln$mu/ln$sig)

hist(lx$tail,breaks=100)
plot(z,lx$tail)
plot(z,log(lx$scale))

hist(nb$size,breaks=100)
```

For neg binom, the size parameter distribution is bimodal with peaks at 0.1 (d2,d4 cell types) and 0.6 (d0,d7 cell types). For poisson-lognormal the sig parameter distribution is bimodal with peaks at 1.1 (d0,d7) and 2.0 (d2,d4). For poisson-lomax the tail parameter distribution is bimodal with peaks at 1.1 (d2,d4) and 3.0 (d0,d7). The bimodality in all cases is determined by the cell type.

```{r}
boxplot(nb$size~cm$cell_type1)
boxplot(ln$sig~cm$cell_type1)
boxplot(lx$tail~cm$cell_type1)
```
