---
title: "Quasi-UMIs for Klein 2015 (indrops)"
author: "Will Townes"
output: html_document
---

Estimate poisson-lomax, poisson-lognormal, and negative binomial (poisson-gamma) parameters for each cell

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
source("./algs/quminorm.R") #also loads nblomax.R
source("./algs/poilog.R")
fp<-file.path
bp<-"./real/klein_2015"
plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)) dir.create(plt_dir,recursive=TRUE)
```

data loading

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-counts(sce)
cm<-colData(sce)
```

## Exploratory log-log plots

PMF of random cell, compare poisson-lomax with poisson-lognormal

```{r}
plt_name<-"lpmf_llcurve_klein.pdf"
i<-1146
colnames(m)[i] #cell ID="cell1146"
x<-m[,i]

(th<-poilog_mle(x))
(th2<-plomax_mle(x,quadpts=1000))
(th3<-nb_mle(x))
res<-lpmf(x,doplot=FALSE)
xmax<-expm1(1.1*max(res$x))
ymin<-1.1*min(res$y)

fpoilog<-function(t){
  sads::dpoilog(floor(exp(t)),mu=th[1],sig=th[2],log=TRUE)
}
fplomax<-function(t){
  dplomax(floor(exp(t)),tail=th2[1],scale=th2[2],log=TRUE,quadpts=5000)
}
fnb<-function(t){
  dnbinom(floor(exp(t)),size=th3[1],mu=th3[2],log=TRUE)
}

pdf(file=fp(plt_dir,plt_name),width=4,height=4)
marpar<-par("mar"); par(mar=c(4,4,.5,.5))
lpmf_xtra(res,xlab="UMI count",ylab="log(density)",connect=FALSE,ylim=c(ymin,0),xlim=c(-log(2),log(xmax)),cex=2,lwd=3)
curve(fpoilog,from=log(1),to=log(xmax),add=TRUE,col="blue",lwd=3)
curve(fplomax,from=log(1),to=log(xmax),add=TRUE,col="green",lwd=3,lty=2)
curve(fnb,from=log(1),to=log(xmax),add=TRUE,col="red",lwd=3,lty=3)
points(rep(-log(2),3),c(fpoilog(-Inf),fplomax(-Inf),fnb(-Inf)),col=c("blue","green","red"),pch=c(1,1,19),cex=c(1.6,.9,.4),lwd=3)
legend("topright",c("Poisson-lognormal","Poisson-Lomax","negative binomial"),col=c("blue","green","red"),lty=c(1,2,3),lwd=3)
par(mar=marpar)
dev.off()
```

## Compute MLEs for all cells 

`time Rscript ./real/util/mles_qumi.R ./real/klein_2015 data/01_sce_all_genes_all_cells.rds` 
or 
`sbatch -J klein ./real/util/mles_qumi.slurm ./real/klein_2015 data/01_sce_all_genes_all_cells.rds`

## Goodness of fit 

### BIC

```{r}
pd<-read.table(fp(bp,"results/mle_merged.txt"),header=TRUE)
ln<-subset(pd,method=="poilog")
lx<-subset(pd,method=="poilomax")
nb<-subset(pd,method=="negbinom")
#when is poilog best?
mean(ln$bic<pmin(lx$bic,nb$bic))
#when is poisson-lomax best?
mean(lx$bic<pmin(ln$bic,nb$bic))
#nb best?
mean(nb$bic<pmin(ln$bic,lx$bic))
with(pd,boxplot(bic~method)) #very small differences
```

Based on BIC, in 60% of cells, Poisson-lognormal fits best. In 40% of cells Poisson-lomax fits best. In 0% of cells, negative binomial fits best. The overall differences between methods are very small according to BIC.

### Predictive check based on simulated maxima

See *real/zhao_2019/02_qumi.Rmd* for a details.

```{r}
cpmax<-read.table(fp(bp,"results/compare_maxima.txt"),header=TRUE)
cpmax$log_err<-log(cpmax$sim_max)-log(cpmax$true_max)
ggplot(cpmax,aes(x=log_err))+geom_histogram()+facet_wrap(~method,scales="free")
```

Negative binomial underestimates the maximum, Poisson-lognormal is well-calibrated, and Poisson-Lomax overestimates.

## Distributions of parameter estimates (MLEs)

```{r}
pd<-read.table(fp(bp,"results/mle_merged.txt"),header=TRUE)
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
sce<-sce[,as.character(unique(pd$cell))]
cm<-colData(sce)
m<-counts(sce)
cm$cell<-rownames(cm)
pd<-merge(pd,as.data.frame(cm),by="cell")
```

```{r}
ggplot(pd,aes(x=shape))+geom_histogram(bins=100)+facet_wrap(~method,scales="free")
ggplot(pd,aes(x=cell_type1,y=shape))+geom_boxplot()+facet_wrap(~method,scales="free")
```

For neg binom, the size parameter distribution is bimodal with peaks at 0.1 (d2,d4 cell types) and 0.6 (d0,d7 cell types). For poisson-lognormal the sig parameter distribution is bimodal with peaks at 1.1 (d0,d7) and 2.0 (d2,d4). For poisson-lomax the tail parameter distribution is bimodal with peaks at 1.1 (d2,d4) and 3.0 (d0,d7). The bimodality in all cases is determined by the cell type.

### Comparing fitted MLEs to summary statistics

```{r}
# cm<-colData(sce)
# z<-log10(colSums(m))
# boxplot(ln$sig~cm$cell_type1)
# boxplot(z~cm$cell_type1)
# plot(z,ln$sig,col=cm$cell_type1)
# plot(z,ln$mu)
# plot(z,ln$mu/ln$sig)
# plot(z,lx$tail)
# plot(z,log(lx$scale))
```

### Compare MLE to MOM estimator

Comparing MLEs of scale parameter to MOM of scale param with fixed shape

```{r}
# lpz<-log(1-colMeans(m>0))
# scale_mom<-poilog_pzero2mu(lpz,sig=2.0)
# pd<-data.frame(scale_mom=scale_mom,celltype=cm$cell_type1,shape_mle=ln$sig,scale_mle=ln$mu)
# ggplot(pd,aes(x=scale_mle,y=scale_mom,colour=shape_mle))+geom_point()+geom_abline(intercept=0,slope=1)+scale_color_gradient2(low="orange",mid="blue",high="red",midpoint = 2.0)+xlab("scale parameter MLE")+ylab("scale parameter MOM estimate")
# ggsave(fp(plt_dir,"klein_mom_vs_mle.pdf"),width=6,height=4)
```