---
title: "Quasi-UMIs for Tung 2016"
author: "Will Townes"
date: "4/10/2019"
output: html_document
---

Exploring reads vs UMIs with goal of normalizing read counts to quasi-UMIs

* Estimate discrete power law distribution for many UMI samples
* Quantile normalize read counts to quasi UMI counts
* Compare qumi distribution to actual UMI distribution

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
library(ggridges)
source("./algs/quminorm.R") #also loads nblomax.R,poilog.R
fp<-file.path
bp<-"./real/tung_2016"

plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)){ dir.create(plt_dir,recursive=TRUE) }
```

data loading

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-counts(sce)
rc<-assay(sce,"read_counts")
```

PMF of random cell, compare poisson-lomax with poisson-lognormal

```{r}
#i<-1
i<-sample.int(ncol(m),size=1)
x<-m[,i]
(th<-poilog_mle(x))
(th2<-plomax_mle(x,quadpts=1000))
(th3<-nb_mle(x))
lpmf(x)
llcurve_poilog(max(x),th,col="blue",lwd=2)
llcurve_lomax(max(x),th2,lik="poi",col="green",lty=2,lwd=2,q=5000)
llcurve_nb(max(x),th3,col="red",lty=3,lwd=2)
legend("topright",c("poisson-lognormal","poisson-lomax","negative binomial"),col=c("blue","green","red"),lty=c(1,2,3),lwd=2)
```

make lpmf plots for all cells

```{r}
f<-function(t){
  df<-lpmf(m[,t],doplot=FALSE)
  df$id<-t
  df
}
res<-do.call(rbind,lapply(seq.int(ncol(m)),f))
ggplot(res,aes(x=x,y=y,group=id))+geom_line(alpha=.1)+xlab("log(1+x)")+ylab("log(density)")
ggsave(fp(plt_dir,"lpmf_allcells_tung.pdf"),width=6,height=4)
```

apply QUMI to entire data matrix 

```{bash}
Rscript ./real/util/quminorm_script.R ./real/tung_2016/data/01_sce_all_genes_all_cells.rds ./real/tung_2016/data/02_sce_qumi.rds
```
or 
```{bash}
sbatch -J tung ./real/util/quminorm_script.slurm ./real/tung_2016/data/01_sce_all_genes_all_cells.rds ./real/tung_2016/data/02_sce_qumi.rds
```

For sensitivity analysis, additional poilog quminorm with sig=1.0 instead of 2.0

```{r}
sce2<-readRDS(fp(bp,"data/02_sce_qumi.rds"))
#1 minute
system.time(assay(sce2,"qumi_poilog1")<-quminorm_matrix(rc,1.0,lik="poilog"))
saveRDS(sce2,file=fp(bp,"data/02_sce_qumi.rds"))
```

Show existing normalizations don't align distributions of technical replicates

```{r}
sce2<-readRDS(fp(bp,"data/02_sce_qumi.rds"))
cm<-colData(sce2)
ss<-sce2[,cm$individual=="NA19098" & cm$replicate=="r3"]
ncell<-20
ss<-ss[,sample.int(ncol(ss),ncell)]
cellnames<-paste0("cell",1:ncell)
res<-list()
for(m in c("read_counts","census_counts","cpm","scran_counts","counts","qumi_poilog")){
  X<-assay(ss,m)
  f<-function(i){
    x<-X[,i]
    data.frame(x=log10(x[x>0]),mth=m,cell=cellnames[i])
  }
  res[[m]]<-do.call(rbind,lapply(seq.int(ncol(ss)),f))
}
pd<-do.call(rbind,res)
levels(pd$mth)[levels(pd$mth)=="counts"] <- "umi_counts"

pd2<-subset(pd,mth %in% c("read_counts","cpm","census_counts"))
pd2$mth<-factor(pd2$mth,levels=c("read_counts","cpm","census_counts"))
ggplot(pd2,aes(x=x,group=cell))+geom_density(alpha=.9)+facet_wrap(~mth,scales="free")+xlab("log10(expression)")
ggsave(fp(bp,"results/fig/normalize_tung_distr.pdf"),width=8,height=3)

#split plots into separate files for latex
ggplot(subset(pd2,mth=="read_counts"),aes(x=x,group=cell))+geom_density(alpha=.9)+xlab("log10(expression)")
ggsave(fp(bp,"results/fig/normalize_tung_distr_rc.pdf"),width=3,height=3)

ggplot(subset(pd2,mth=="cpm"),aes(x=x,group=cell))+geom_density(alpha=.9)+xlab("log10(expression)")
ggsave(fp(bp,"results/fig/normalize_tung_distr_cpm.pdf"),width=3,height=3)

ggplot(subset(pd2,mth=="census_counts"),aes(x=x,group=cell))+geom_density(alpha=.9)+xlab("log10(expression)")
ggsave(fp(bp,"results/fig/normalize_tung_distr_census.pdf"),width=3,height=3)
```

compare distances between normalizations for all cells. Using the UMI counts as a gold standard, we compute the euclidean distance between nonzero normalized counts and the UMI counts on the log scale for each normalization.

```{r}
#sce2<-readRDS(fp(bp,"data/02_sce_qumi.rds"))
res<-distance_compare(sce2)
#sce3<-sce2
#assays(sce3)<-assays(sce2)[c("counts","qumi_poilog1")]
#res2<-distance_compare(sce3)
#res$qumi_poilog1<-res2$qumi_poilog1
write.table(res,file=fp(bp,"results/qumi_distance.txt"),quote=FALSE,row.names=FALSE)
```

```{r}
#i<-sample.int(ncol(rc),size=1)
umi<-m[,i]
x<-rc[,i]
#poisson-lomax qumi
(th<-plomax_mle(umi,quadpts=1000))
xqumi<-quminorm_plomax(x,shape=th[1],sc=th[2])
#poisson-lognormal qumi
(th2<-poilog_mle(umi))
xqumi2<-quminorm_poilog(x,shape=th2[2],sc=th2[1])
#negative binomial qumi
(th3<-nb_mle(umi))
xqumi3<-quminorm_nb(x,shape=th3[1],sc=th3[2])
#plot qumi distributions and compare to actual UMI
lpmf(xqumi,type="l",col="green",lty=2,lwd=2)
lpmf(umi,add=TRUE,type="p",cex=2)
lpmf(xqumi2,add=TRUE,type="l",col="blue",lty=1,lwd=2)
lpmf(xqumi3,add=TRUE,type="l",col="red",lty=3,lwd=2)
cc<-ceiling(assay(sce2,"census_counts")[,i])
lpmf(cc,add=TRUE,type="l",col="orange",lty=2,lwd=2)
legend("topright",c("negative binomial","poisson-lognormal","poisson-lomax","census counts"),lty=c(3,1,2,2),col=c("red","blue","green","orange"),lwd=2)
title("Quasi-UMIs vs Actual UMIs")
```

```{r}
#Poisson-lomax tends to overestimate
plot(log1p(umi),log1p(xqumi),main="Poisson-Lomax Quasi-UMIs")
abline(0,1)
#neg binom tends to underestimate
plot(log1p(umi),log1p(xqumi3),main="Negative Binomial Quasi-UMIs")
abline(0,1)
#Poisson-lognormal does well
plot(log1p(umi),log1p(xqumi2),main="Poisson-Lognormal Quasi-UMIs")
abline(0,1)
#census counts
plot(log1p(umi),log1p(cc),main="Census Counts")
abline(0,1)
```

#see if quasi-UMI improves over other normalizations

graphical comparison of methods. Qumis are nearly always closer to the true UMI counts than read counts or CPM.

```{r}
library(tidyverse)
res<-read.table(fp(bp,"results/qumi_distance.txt"),header=TRUE,stringsAsFactors=FALSE)
pd<-tidyr::gather(res,key="method",value="distance",-cell)
pd<-subset(pd,method!="counts")
pd2<-pd %>% group_by(method) %>% summarize(med_dist=median(distance))
o<-order(pd2$med_dist)
pd$method<-factor(pd$method,levels=pd2$method[o])
ggplot(pd,aes(x=method,y=distance))+geom_boxplot()+scale_y_log10()+theme(axis.text.x = element_text(angle = 15, hjust=1))+ylab("distance from UMI counts")
ggsave(fp(plt_dir,"dist_comp_tung.pdf"),width=6,height=4)
```
