---
title: "Quasi-UMIs for Tung 2016 iPSCs"
author: "Will Townes"
output: html_document
---

Exploring reads vs UMIs with goal of normalizing read counts to quasi-UMIs

* Estimate discrete power law distribution for many UMI samples
* Quantile normalize read counts to quasi UMI counts
* Compare qumi distribution to actual UMI distribution

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
library(ggridges)
source("algs/quminorm.R") #also loads nblomax.R,poilog.R
fp<-file.path
bp<-"real/tung_2016"

plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)){ dir.create(plt_dir,recursive=TRUE) }
```

data loading

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-counts(sce)
rc<-assay(sce,"read_counts")
```

PMF of random cell, compare poisson-lomax with poisson-lognormal

```{r}
#i<-1
i<-sample.int(ncol(m),size=1)
x<-m[,i]
(th<-poilog_mle(x))
(th2<-plomax_mle(x,quadpts=1000))
(th3<-nb_mle(x))
lpmf(x)
llcurve_poilog(max(x),th,col="blue",lwd=2)
llcurve_lomax(max(x),th2,lik="poi",col="green",lty=2,lwd=2,q=5000)
llcurve_nb(max(x),th3,col="red",lty=3,lwd=2)
legend("topright",c("poisson-lognormal","poisson-lomax","negative binomial"),col=c("blue","green","red"),lty=c(1,2,3),lwd=2)
```

make lpmf plots for all cells

```{r}
f<-function(t){
  df<-lpmf(m[,t],doplot=FALSE)
  df$id<-t
  df
}
res<-do.call(rbind,lapply(seq.int(ncol(m)),f))
ggplot(res,aes(x=x,y=y,group=id))+geom_line(alpha=.1)+xlab("log(1+x)")+ylab("log(density)")
ggsave(fp(plt_dir,"lpmf_allcells_tung.pdf"),width=6,height=4)
```

### Apply QUMI to entire data matrix 

First we create a configuration file (JSON) to keep track of all the shape parameters for each model. Note that for Poisson-lognormal, 2.4 is the median of the iPSC-specific MLE distribution in the *schiebinger_2019* training dataset.

```{r}
shp<-list(nb=.1,poilog=c(1,2,2.4,3),plomax=1)
write(jsonlite::toJSON(shp,pretty=TRUE),file=fp(bp,"quminorm_script_config.json"))
```

```{bash}
Rscript ./real/util/quminorm_script.R ./real/tung_2016 data/01_sce_all_genes_all_cells.rds data/02_sce_qumi.rds
```
or 
```{bash}
sbatch -J tung ./real/util/quminorm_script.slurm ./real/tung_2016 data/01_sce_all_genes_all_cells.rds data/02_sce_qumi.rds
```

Show existing normalizations don't align distributions of technical replicates

```{r}
set.seed(200)
sce2<-readRDS(fp(bp,"data/02_sce_qumi.rds"))
cm<-as.data.frame(colData(sce2))
ss<-sce2[,cm$individual=="NA19098" & cm$replicate=="r3"]
ncell<-20
ss<-ss[,sample.int(ncol(ss),ncell)]
cellnames<-paste0("cell",1:ncell)
res<-list()
for(m in c("read_counts","census_counts","cpm","scran_counts","counts","qumi_poilog_2.4")){
  X<-assay(ss,m)
  f<-function(i){
    x<-X[,i]
    data.frame(x=log10(x[x>0]),mth=m,cell=cellnames[i])
  }
  res[[m]]<-do.call(rbind,lapply(seq.int(ncol(ss)),f))
}
pd<-do.call(rbind,res)
pd$mth<-factor(pd$mth)
levels(pd$mth)[levels(pd$mth)=="counts"] <- "umi_counts"
levels(pd$mth)[levels(pd$mth)=="qumi_poilog_2.4"] <- "qumi_counts"
lvl<-c("read_counts","umi_counts","qumi_counts","cpm","scran_counts","census_counts")
pd$mth<-factor(pd$mth,levels=lvl)

ggplot(pd,aes(x=x,group=cell))+geom_density(alpha=.9)+facet_wrap(~mth,scales="free",ncol=3)+xlab("log10(expression)")
ggsave(fp(plt_dir,"normalize_tung_distr.pdf"),width=7,height=5)

#split plots into separate files for latex
plt_func<-function(m){
  ggplot(subset(pd,mth==m),aes(x=x,group=cell))+geom_density(alpha=.9)+xlab("log10(expression)")
}
plt<-plt_func("read_counts")
ggsave(fp(plt_dir,"normalize_tung_distr_rc.pdf"),plot=plt,width=3,height=3)
plt<-plt_func("umi_counts")
ggsave(fp(plt_dir,"normalize_tung_distr_umi.pdf"),plot=plt,width=3,height=3)
plt<-plt_func("qumi_counts")
ggsave(fp(plt_dir,"normalize_tung_distr_qumi.pdf"),plot=plt,width=3,height=3)
plt<-plt_func("cpm")
ggsave(fp(plt_dir,"normalize_tung_distr_cpm.pdf"),plot=plt,width=3,height=3)
plt<-plt_func("scran_counts")
ggsave(fp(plt_dir,"normalize_tung_distr_scran.pdf"),plot=plt,width=3,height=3)
plt<-plt_func("census_counts")
ggsave(fp(plt_dir,"normalize_tung_distr_census.pdf"),plot=plt,width=3,height=3)
```

compare distances between normalizations for all cells. Using the UMI counts as a gold standard, we compute the euclidean distance between nonzero normalized counts and the UMI counts on the log scale for each normalization.

```{r}
#sce2<-readRDS(fp(bp,"data/02_sce_qumi.rds"))
system.time(res<-distance_compare(sce2))
write.table(res,file=fp(bp,"results/qumi_distance.txt"),quote=FALSE,row.names=FALSE)
```
or
```{bash}
Rscript ./real/util/qumi_dist_compare_script.R ./real/tung_2016/data/02_sce_qumi.rds ./real/tung_2016/results/qumi_distance.txt
```
or
```{bash}
sbatch -J tung_dist_comp ./real/util/qumi_dist_compare_script.slurm ./real/tung_2016/data/02_sce_qumi.rds ./real/tung_2016/results/qumi_distance.txt
```

compare distances for relative abundances

```{r}
rafunc<-function(m){
  t(t(m)/Matrix::colSums(m))
}
sce3<-sce2
for(a in assayNames(sce3)){
  if(a %in% c("cpm","scran_counts")){
    assay(sce3,a)<-NULL
  } else {
    assay(sce3,a)<-rafunc(assay(sce3,a))
  }
}
res3<-distance_compare(sce3)
write.table(res3,file=fp(bp,"results/qumi_ra_distance.txt"),quote=FALSE,row.names=FALSE)
```

### see if quasi-UMI improves over other normalizations

graphical comparison of methods. Qumis are nearly always closer to the true UMI counts than read counts or CPM.

```{r}
library(tidyverse)
res<-read.table(fp(bp,"results/qumi_distance.txt"),header=TRUE,stringsAsFactors=FALSE)
pd<-tidyr::gather(res,key="method",value="distance",-cell)
pd<-subset(pd,method!="counts")
pd2<-pd %>% group_by(method) %>% summarize(med_dist=median(distance))
o<-order(pd2$med_dist)
pd$method<-factor(pd$method,levels=pd2$method[o])
ggplot(pd,aes(x=method,y=distance))+geom_boxplot()+scale_y_log10()+theme(axis.text.x = element_text(angle = 15, hjust=1))+ylab("distance from UMI counts")
ggsave(fp(plt_dir,"dist_comp_tung.pdf"),width=6,height=4)
```

alternative approach, compare relative abundances instead of counts

```{r}
res<-read.table(fp(bp,"results/qumi_ra_distance.txt"),header=TRUE,stringsAsFactors=FALSE)
pd<-tidyr::gather(res,key="method",value="distance",-cell)
pd<-subset(pd,method!="counts")
pd2<-pd %>% group_by(method) %>% summarize(med_dist=median(distance))
o<-order(pd2$med_dist)
pd$method<-factor(pd$method,levels=pd2$method[o])
ggplot(pd,aes(x=method,y=distance))+geom_boxplot()+scale_y_log10()+theme(axis.text.x = element_text(angle = 15, hjust=1))+ylab("distance from UMI counts rel. abund.")
#ggsave(fp(plt_dir,"dist_comp_ra_tung.pdf"),width=6,height=4)
```

### MA plot and CV comparison between normalizations

A reviewer suggests that the differences between QUMIs and UMIs may depend on the expression level of the gene. To visualize this, we compute an MA plot between the UMI counts and each of the normalizations. Since MA plot requires a single vector for each "condition", we just average the counts across all cells for each gene.

We also compute the MA plot contrasting the coefficients of variation instead of the average expression.

We repeat both MA plot comparisons using relative abundances instead of counts.

```{r}
sce<-readRDS(fp(bp,"data/02_sce_qumi.rds"))
umi<-assay(sce,"counts")
nz_ct<-rowSums(umi>0)
nz_pct<-nz_ct/ncol(umi)
x<-log2(rowMeans(umi))
cv0<-log2(apply(umi,1,sd))-x
umi_ra<-t(t(umi)/colSums(umi))
x_ra<-log2(rowMeans(umi_ra))
cv0_ra<-log2(apply(umi_ra,1,sd))-x_ra
f<-function(a){
  alt<-assay(sce,a)
  y<-log2(rowMeans(alt))
  cv<-log2(apply(alt,1,sd))-y
  alt_ra<-t(t(alt)/colSums(alt))
  y_ra<-log2(rowMeans(alt_ra))
  cv_ra<-log2(apply(alt_ra,1,sd))-y_ra
  data.frame(method=a, nz_ct=nz_ct, nz_pct=nz_pct,
             mean_a=.5*(x+y), mean_m=y-x,
             cv_a=.5*(cv+cv0), cv_m=cv-cv0,
             ra_mean_a=.5*(x_ra+y_ra), ra_mean_m=y_ra-x_ra,
             ra_cv_a=.5*(cv_ra+cv0_ra), ra_cv_m=cv_ra-cv0_ra)
}
anames<-c("read_counts","qumi_poilog_2","scran_counts","census_counts")
res<-do.call(rbind,lapply(anames,f))
res$method<-plyr::mapvalues(res$method,from="qumi_poilog_2",to="qumi_counts")
write.table(res,fp(bp,"results/qumi_maplot.txt"),quote=FALSE,row.names=FALSE)
#rownames(res)<-NULL
ggplot(res,aes(x=mean_a,y=mean_m))+geom_point(alpha=.9,size=.5)+facet_wrap(~method,nrow=2,scales="free_x")+xlab("average log expression (A)")+ylab("difference in log expression (M)")+geom_hline(yintercept=0,colour="blue")
ggplot(res,aes(x=cv_a,y=cv_m))+geom_point(alpha=.9,size=.5)+facet_wrap(~method,nrow=2,scales="free_x")+xlab("average log CV (A)")+ylab("difference in log CV (M)")+geom_hline(yintercept=0,colour="blue")
ggplot(res,aes(x=ra_mean_a,y=ra_mean_m))+geom_point(alpha=.9,size=.5)+facet_wrap(~method,nrow=2,scales="free_x")+xlab("average log expression (A)")+ylab("difference in log expression (M)")+geom_hline(yintercept=0,colour="blue")
ggplot(res,aes(x=ra_cv_a,y=ra_cv_m))+geom_point(alpha=.9,size=.5)+facet_wrap(~method,nrow=2,scales="free_x")+xlab("average log CV (A)")+ylab("difference in log CV (M)")+geom_hline(yintercept=0,colour="blue")
```
