---
title: "Quasi-UMIs for Cao 2017"
author: "Will Townes"
output: html_document
---

Estimate poisson-lomax, poisson-lognormal, and negative binomial (poisson-gamma) parameters for each cell

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
source("./algs/quminorm.R") #also loads nblomax.R, poilog.R
fp<-file.path
bp<-"./real/cao_2017"
plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)){dir.create(plt_dir,recursive=TRUE)}
```

data loading

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-counts(sce)
```

## Exploratory log-log plots

PMF of random cell, compare poisson-lomax with poisson-lognormal

```{r}
i<-6272 #cell ID="cele-002-090.AATCATACGG"
#i<-sample.int(ncol(m),size=1)
x<-m[,i]
max(x) #26

xt<-table(x)
xt2<-xt/sum(xt)
xpts<-as.integer(names(xt2))
txtlabs<-format(xt,big.mark=",",trim=TRUE)

#compare histogram to lpmf
pdf(file=fp(plt_dir,"cao_a_hist.pdf"),width=4,height=4)
plot(xt2,lwd=5,ylim=c(0,1.1),xlim=c(-1,max(x)),xlab="UMI count",ylab="fraction of genes")
txty<-xt2+0.05
txty[2:5]<-txty[2:5]*c(5,4,3,2)
text(xpts,y=txty,labels=txtlabs,adj=0.3,cex=0.75)
dev.off()

pdf(file=fp(plt_dir,"cao_b_hist_logy.pdf"),width=4,height=4)
ypts<-log10(as.numeric(xt2))
plot(xpts,ypts,type="b",ylim=c(min(ypts),0.5),xlim=c(-1,max(x)),xlab="UMI count",ylab="log10(fraction of genes)")
text(xpts,ypts+0.3,labels=txtlabs,adj=0,cex=0.75)
dev.off()

pdf(file=fp(plt_dir,"cao_c_lpmf.pdf"),width=4,height=4)
res<-lpmf(x,doplot=FALSE,logbase=10)
plot(res$x,res$y,xlab="log10(1+UMI count)",ylab="log10(density)",type="b")
dev.off()

#larger lpmf on log10 scale not log10(1+...)
res<-lpmf(x,doplot=FALSE)
pdf(file=fp(plt_dir,"cao_lpmf.pdf"),width=4,height=4)
lpmf_xtra(res,xlab="UMI count",ylab="log(density)")
dev.off()
```

make lpmf plots for all cells

```{r}
f<-function(t){
  df<-lpmf(m[,t],doplot=FALSE)
  df$id<-t
  df
}
col_ids<-sample(ncol(m),size=500)
res<-do.call(rbind,lapply(col_ids,f)) #slow
res$x<-expm1(res$x)
res$lx<-log(res$x)
res$lx[is.infinite(res$lx)]<- -log(2)
res_z<-subset(res,x==0)
res_nz<-subset(res,x>0)

ggplot(res)+geom_line(data=subset(res,x>0),aes(x=x,y=y,group=id),alpha=.2)+xlab("UMI count")+ylab("log(density)")+geom_point(data=subset(res,x==0),aes(x=1/2,y=y))+scale_x_log10(limits=c(1/2,NA),breaks=c(1/2,1,10,100),labels=c(0,1,10,100))+ylim(NA,max(res$y))+geom_vline(xintercept=sqrt(.5),color="#CCCCCC",size=10)
ggsave(fp(plt_dir,"lpmf_500cells_cao.pdf"),width=6,height=4)
```

lpmf vs MLEs

```{r}
plt_name<-"lpmf_llcurve_cao.pdf"
i<-6272 #cell ID="cele-002-090.AATCATACGG"
#set.seed(222)
#i<-sample.int(ncol(m),size=1)
x<-m[,i]

(th<-poilog_mle(x))
(th2<-plomax_mle(x,quadpts=1000))
(th3<-nb_mle(x))
res<-lpmf(x,doplot=FALSE)
xmax<-expm1(1.1*max(res$x))
ymin<-1.1*min(res$y)

fpoilog<-function(t){
  sads::dpoilog(floor(exp(t)),mu=th[1],sig=th[2],log=TRUE)
}
fplomax<-function(t){
  dplomax(floor(exp(t)),tail=th2[1],scale=th2[2],log=TRUE,quadpts=5000)
}
fnb<-function(t){
  dnbinom(floor(exp(t)),size=th3[1],mu=th3[2],log=TRUE)
}

pdf(file=fp(plt_dir,plt_name),width=4,height=4)
marpar<-par("mar"); par(mar=c(4,4,.5,.5))
lpmf_xtra(res,xlab="UMI count",ylab="log(density)",connect=FALSE,ylim=c(ymin,0),xlim=c(-log(2),log(xmax)),cex=2,lwd=3)
curve(fpoilog,from=log(1),to=log(xmax),add=TRUE,col="blue",lwd=3)
curve(fplomax,from=log(1),to=log(xmax),add=TRUE,col="green",lwd=3,lty=2)
curve(fnb,from=log(1),to=log(xmax),add=TRUE,col="red",lwd=3,lty=3)
points(rep(-log(2),3),c(fpoilog(-Inf),fplomax(-Inf),fnb(-Inf)),col=c("blue","green","red"),pch=c(1,1,19),cex=c(1.6,.9,.4),lwd=3)
legend("topright",c("Poisson-lognormal","Poisson-Lomax","negative binomial"),col=c("blue","green","red"),lty=c(1,2,3),lwd=3)
par(mar=marpar)
dev.off()
```

## Compute MLEs for all cells 

`time Rscript ./real/util/mles_qumi.R ./real/cao_2017 data/01_sce_all_genes_all_cells.rds` 
or 
`sbatch -J cao ./real/util/mles_qumi.slurm ./real/cao_2017 data/01_sce_all_genes_all_cells.rds`

## Goodness of fit 

### BIC

```{r}
pd<-read.table(fp(bp,"results/mle_merged.txt"),header=TRUE)
ln<-subset(pd,method=="poilog")
lx<-subset(pd,method=="poilomax")
nb<-subset(pd,method=="negbinom")
#when is poilog best?
mean(ln$bic<pmin(lx$bic,nb$bic))
#when is poisson-lomax best?
mean(lx$bic<pmin(ln$bic,nb$bic))
#nb best?
mean(nb$bic<pmin(ln$bic,lx$bic))
with(pd,boxplot(bic~method)) #very small differences
```

Based on BIC, in 90% of cells, negative binomial fits best. In 5% of cells Poisson-lomax fits best. In 5% of cells, Poisson-lognormal fits best. The overall differences between methods are very small according to BIC.

### Predictive check based on simulated maxima

See *real/zhao_2019/02_qumi.Rmd* for details.

```{r}
cpmax<-read.table(fp(bp,"results/compare_maxima.txt"),header=TRUE)
cpmax$log_err<-log(cpmax$sim_max)-log(cpmax$true_max)
ggplot(cpmax,aes(x=log_err))+geom_histogram()+facet_wrap(~method,scales="free")
```

Negative binomial underestimates the maximum, Poisson-lognormal is well-calibrated, and Poisson-Lomax overestimates.

## Distributions of parameter estimates (MLEs)

```{r}
pd<-read.table(fp(bp,"results/mle_merged.txt"),header=TRUE)
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
sce<-sce[,as.character(unique(pd$cell))]
cm<-colData(sce)
m<-counts(sce)
cm$cell<-rownames(cm)
pd<-merge(pd,as.data.frame(cm),by="cell")
```

```{r}
ggplot(pd,aes(x=shape))+geom_histogram(bins=100)+facet_wrap(~method,scales="free")+scale_x_log10()
ggplot(pd,aes(x=tissue,y=shape))+geom_boxplot()+facet_grid(~method,scales="free_x")+scale_y_log10()+coord_flip()
ggplot(pd,aes(x=cell.type,y=shape))+geom_boxplot()+facet_grid(~method,scales="free_x")+scale_y_log10()+coord_flip()
```

For neg binom, the size parameter ranges from .003 to .1 with .015 being typical. For poisson-lognormal the sig parameter ranges from 2-3 with 2.5 typical. For poisson-lomax the tail parameter is around 0.7-0.8. 

There does appear to be some tissue and cell-type specificity to the shape parameter estimates.

### Comparing fitted MLEs to summary statistics

```{r}
# z<-log10(Matrix::colSums(m))
# plot(z,ln$sig)
# plot(z,ln$mu)
# plot(z,ln$mu/ln$sig)
# plot(z,lx$tail,ylim=c(0,2))
# plot(z,log(lx$scale))
```

### Compare MLE to MOM estimator

infer params from prob zero

```{r}
# lpz<-log(1-Matrix::colMeans(m>0))
# nb_mu2<-nb_pzero2mu(lpz,nb$size)
# plot(nb$mu,nb_mu2) #excellent agreement
# 
# ln_mu2<-poilog_pzero2mu(lpz,ln$sig)
# plot(ln$mu,ln_mu2) #excellent agreement
```

## Stuff that didn't work or wasn't used

```{r}
# #vuong test
# i<-1
# plg_lp<-function(x,th){sads::dpoilog(x,mu=th[1],sig=th[2],log=TRUE)}
# nb_lp<-function(x,th){dnbinom(x,size=th[1],mu=th[2],log=TRUE)}
# plx_lp<-function(x,th){dplomax(x,tail=th[1],scale=th[2],log=TRUE,quadpts=5000)}
# lpfuncs<-list(nb=nb_lp,plg=plg_lp,plx=plx_lp)
# 
# y<-m[,i]
# th1<-as.numeric(ln[i,c("mu","sig")])
# th2<-as.numeric(nb[i,c("size","mu")])
# lpfunc1<-plg_lp; lpfunc2<-nb_lp
# 
# xt<-table(y)
# x<-as.integer(names(xt))
# n<-as.integer(xt)
# 
# vuong_discrete<-function(x,n,lpfunc1,lpfunc2,th1,th2){
#   #test whether model for lpfunc1 is better than lpfunc2 for data y
#   #x=unique values of data
#   #n=number of times each value was observed
#   #can get x,n from table(y)
#   #lpfunc1,lpfunc2=some functions of x giving log-likelihoods
#   #th1,th2=parameters of lpfunc1,lpfunc2
#   delta<-lpfunc1(x,th1)-lpfunc2(x,th2)
#   s<-sqrt(Hmisc::wtd.var(delta,n)/sum(n))
#   z<-Hmisc::wtd.mean(delta,n)/s
#   pval<-2*pnorm(abs(z),lower.tail=FALSE)
#   c(zstat=z,pval=pval)
# }
```
