---
title: "Quasi-UMIs for Cao 2017"
author: "Will Townes"
output: html_document
---

Estimate poisson-lomax, poisson-lognormal, and negative binomial (poisson-gamma) parameters for each cell

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
source("./algs/quminorm.R") #also loads nblomax.R, poilog.R
fp<-file.path
bp<-"./real/cao_2017"
plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)){dir.create(plt_dir,recursive=TRUE)}
```

data loading

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-counts(sce)
```

PMF of random cell, compare poisson-lomax with poisson-lognormal

```{r}
i<-6272 #cell ID="cele-002-090.AATCATACGG"
#i<-sample.int(ncol(m),size=1)
x<-m[,i]
max(x) #26

xt<-table(x)
xt2<-xt/sum(xt)
xpts<-as.integer(names(xt2))
txtlabs<-format(xt,big.mark=",",trim=TRUE)

#compare histogram to lpmf
pdf(file=fp(plt_dir,"cao_a_hist.pdf"),width=4,height=4)
plot(xt2,lwd=5,ylim=c(0,1.1),xlim=c(-1,max(x)),xlab="UMI count",ylab="fraction of genes")
txty<-xt2+0.05
txty[2:5]<-txty[2:5]*c(5,4,3,2)
text(xpts,y=txty,labels=txtlabs,adj=0.3,cex=0.75)
dev.off()

pdf(file=fp(plt_dir,"cao_b_hist_logy.pdf"),width=4,height=4)
ypts<-log10(as.numeric(xt2))
plot(xpts,ypts,type="b",ylim=c(min(ypts),0.5),xlim=c(-1,max(x)),xlab="UMI count",ylab="log10(fraction of genes)")
text(xpts,ypts+0.3,labels=txtlabs,adj=0,cex=0.75)
dev.off()

pdf(file=fp(plt_dir,"cao_c_lpmf.pdf"),width=4,height=4)
res<-lpmf(x,doplot=FALSE,logbase=10)
plot(res$x,res$y,xlab="log10(1+UMI count)",ylab="log10(density)",type="b")
dev.off()

#larger lpmf on log10 scale not log10(1+...)
res<-lpmf(x,doplot=FALSE)
pdf(file=fp(plt_dir,"cao_lpmf.pdf"),width=4,height=4)
lpmf_xtra(res,xlab="UMI count",ylab="log(density)")
dev.off()
```

make lpmf plots for all cells

```{r}
f<-function(t){
  df<-lpmf(m[,t],doplot=FALSE)
  df$id<-t
  df
}
col_ids<-sample(ncol(m),size=500)
res<-do.call(rbind,lapply(col_ids,f)) #slow
res$x<-expm1(res$x)
res$lx<-log(res$x)
res$lx[is.infinite(res$lx)]<- -log(2)
res_z<-subset(res,x==0)
res_nz<-subset(res,x>0)

ggplot(res)+geom_line(data=subset(res,x>0),aes(x=x,y=y,group=id),alpha=.2)+xlab("UMI count")+ylab("log(density)")+geom_point(data=subset(res,x==0),aes(x=1/2,y=y))+scale_x_log10(limits=c(1/2,NA),breaks=c(1/2,1,10,100),labels=c(0,1,10,100))+ylim(NA,max(res$y))+geom_vline(xintercept=sqrt(.5),color="#CCCCCC",size=10)
ggsave(fp(plt_dir,"lpmf_500cells_cao.pdf"),width=6,height=4)
```

lpmf vs MLEs

```{r}
i<-6272 #cell ID="cele-002-090.AATCATACGG"
#set.seed(222)
#i<-sample.int(ncol(m),size=1)
x<-m[,i]
(th<-poilog_mle(x))
(th2<-plomax_mle(x,quadpts=1000))
#(th3<-nb_mle(x))
res<-lpmf(x,doplot=FALSE)
xmax<-expm1(4)

fpoilog<-function(t){
  sads::dpoilog(floor(exp(t)),mu=th[1],sig=th[2],log=TRUE)
}
fplomax<-function(t){
  dplomax(floor(exp(t)),tail=th2[1],scale=th2[2],log=TRUE,quadpts=5000)
}

pdf(file=fp(plt_dir,"lpmf_llcurve_cao.pdf"),width=5,height=5)
lpmf_xtra(res,xlab="UMI count",ylab="log(density)",connect=FALSE,ylim=c(-15,0),xlim=c(-log(2),log(xmax)),cex=1.2)
curve(fpoilog,from=log(1),to=log(xmax),add=TRUE,col="blue",lwd=2)
curve(fplomax,from=log(1),to=log(xmax),add=TRUE,col="green",lwd=2,lty=2)
points(rep(-log(2),2),c(fpoilog(-Inf),fplomax(-Inf)),col=c("blue","green"),pch=c(1,19),cex=c(.8,.3))
legend("topright",c("Poisson-lognormal","Poisson-Lomax"),col=c("blue","green"),lty=c(1,2),lwd=2)
dev.off()
```

Compute MLEs for all cells `time Rscript ./real/util/mles_qumi.R ./real/cao_2017` or `sbatch -J cao ./real/util/mles_qumi.slurm ./real/cao_2017`

```{r}
nb<-read.table(fp(bp,"data/mle_nb.txt"),header=TRUE)
lx<-read.table(fp(bp,"data/mle_poilomax.txt"),header=TRUE)
ln<-read.table(fp(bp,"data/mle_poilog.txt"),header=TRUE)
good<-!is.na(lx$tail) & !is.na(ln$mu) & !is.na(nb$size)
mean(good)
nb2<-nb[good,]
lx2<-lx[good,]
ln2<-ln[good,]
#when is poilog best?
mean(ln2$bic<pmin(lx2$bic,nb2$bic))
#when is poisson-lomax best?
mean(lx2$bic<pmin(ln2$bic,nb2$bic))
#nb best?
mean(nb2$bic<pmin(ln2$bic,lx2$bic))
```

Based on BIC, in 90% of cells, negative binomial fits best. In 5% of cells Poisson-lomax fits best. In 5% of cells, Poisson-lognormal fits best.

Comparing fitted MLEs to summary statistics

```{r}
cm<-colData(sce)
z<-log10(Matrix::colSums(m))
hist(ln$sig,breaks=100)
boxplot(ln$sig~cm$tissue)
boxplot(ln$sig~cm$cell.type)
plot(z,ln$sig)
plot(z,ln$mu)
plot(z,ln$mu/ln$sig)

plot(z,lx$tail,ylim=c(0,2))
plot(z,log(lx$scale))
```

For neg binom, the size parameter ranges from .003 to .1 with .015 being typical. For poisson-lognormal the sig parameter ranges from 2-3 with 2.5 typical. For poisson-lomax the tail parameter is around 0.7-0.8.

infer params from prob zero

```{r}
lpz<-log(1-Matrix::colMeans(m>0))
nb_mu2<-nb_pzero2mu(lpz,nb$size)
plot(nb$mu,nb_mu2) #excellent agreement

ln_mu2<-poilog_pzero2mu(lpz,ln$sig)
plot(ln$mu,ln_mu2) #excellent agreement
```

```{r}
#vuong test
i<-1
plg_lp<-function(x,th){sads::dpoilog(x,mu=th[1],sig=th[2],log=TRUE)}
nb_lp<-function(x,th){dnbinom(x,size=th[1],mu=th[2],log=TRUE)}
plx_lp<-function(x,th){dplomax(x,tail=th[1],scale=th[2],log=TRUE,quadpts=5000)}
lpfuncs<-list(nb=nb_lp,plg=plg_lp,plx=plx_lp)

y<-m[,i]
th1<-as.numeric(ln[i,c("mu","sig")])
th2<-as.numeric(nb[i,c("size","mu")])
lpfunc1<-plg_lp; lpfunc2<-nb_lp

xt<-table(y)
x<-as.integer(names(xt))
n<-as.integer(xt)

vuong_discrete<-function(x,n,lpfunc1,lpfunc2,th1,th2){
  #test whether model for lpfunc1 is better than lpfunc2 for data y
  #x=unique values of data
  #n=number of times each value was observed
  #can get x,n from table(y)
  #lpfunc1,lpfunc2=some functions of x giving log-likelihoods
  #th1,th2=parameters of lpfunc1,lpfunc2
  delta<-lpfunc1(x,th1)-lpfunc2(x,th2)
  s<-sqrt(Hmisc::wtd.var(delta,n)/sum(n))
  z<-Hmisc::wtd.mean(delta,n)/s
  pval<-2*pnorm(abs(z),lower.tail=FALSE)
  c(zstat=z,pval=pval)
}
```
